---
title: "Remote Sensing of Biodiversity"
format: html
editor: visual
---

## Introduction

This tutorial is a soft introduction to remote sensing of biodiversity using hyperspectral imagery. The example is exploring our ability to map two large shrub species (_Mimetes fimbriifolius_ and _Leucospermum conocarpodendron_) in the Silvermine section of Table Mountain National Park, South Africa, using AVIRIS-NG data. 

Fires:
- Peck's Balley to the North East, 11th November 2023
- Small patch of Mimete's Valley, 1st December 2024
- Most of Silvermine - 28th April 2025

The analyses use the `terra` and `luna` package to process and analyze hyperspectral remote sensing data for biodiversity applications. We will be working with AVIRIS-NG data from the Silvermine area in the Cape Peninsula, South Africa, collected as part of the BioSCape Campaign. The tutorial will cover data loading, preprocessing, and basic analysis techniques to extract meaningful information about biodiversity from the hyperspectral imagery.


## Setup

Install and call the libraries we need. Note that library(luna) was not on CRAN at the time of writing, so you need to install it from the R Spatial Universe repository. You only need to do this once, but you will need to call `library(luna)` every time you start a new R session.

```{r}
# install.packages('luna', repos='https://rspatial.r-universe.dev')

library(tidyverse) # For general data handling
library(broom) # For tidying model outputs
library(terra) # For handling raster and vector data
library(luna) # A new companion package for terra, but specifically designed for remote sensing analyses
library(mapview) # A useful package for interactive spatial data visualization
library(GGally) # For visualizing relationships between variables
```


## Data description

The data used in this tutorial is from the Airborne Visible/Infrared Imaging Spectrometer - Next Generation (AVIRIS-NG) sensor, which provides high-resolution hyperspectral imagery (imaging spectroscopy). The specific dataset we are using is from the Silvermine area in the Cape Peninsula, South Africa, collected as part of the BioSCape Campaign (www.bioscape.io) on the 26th November 2023. The ground sample distance (spatial resolution) is 2m. The AVIRIS data are provided in NetCDF format and can be accessed [here]() (±600MB). We also have an Area of Interest (AOI) defined as a GeoJSON file and a set of points for extraction in KML format.

## User inputs 

Here we'll set the file paths to where the data are stored. You may need to change these paths to match where you have stored the data on your computer. The `aviris` variable points to the AVIRIS-NG NetCDF file, which is very large. You could download it and save it to your GIT repository, but you'd need to tell GIT to ignore it (add the file path and name to `.gitignore`). Alternatively, you can download the data and save it somewhere on your computer, and then update the `aviris` variable to point to that location. The `aoi_geojson` variable points to the GeoJSON file that defines our area of interest, and the `points_file` variable points to the KML file that contains the points for extraction.

```{r}
aviris_nc <- "/Users/jasper/Documents/Datasets/BioSCape/ANG/CapePeninsula/SilvermineEast/ang20231126t085417_012_L2A_OE_0b4f48b4_RFL_ORT_QL/ang20231126t085417_012_L2A_OE_0b4f48b4_RFL_ORT.nc"
aoi_geojson <- "data/Silvermine_Hons_2026.geojson"
points_file <- "data/SilvermineEM.kml" # points for extraction
```

## Read and crop raster to Area of Interest (AOI)

Here we read in the AVIRIS data and the AOI. Note that I reproject the AOI to match the CRS of the AVIRIS data, which is important for spatial analyses. The `vect()` function from the `terra` package is used to read in the AOI as a vector object, and the `project()` function is used to reproject it to match the CRS of the AVIRIS data. 

> Why not reproject the AVIRIS data instead? Because it's a very large raster and reprojecting it can be computationally intensive, whereas reprojecting the AOI (which is a smaller vector file) is much faster.

```{r}
aviris <- rast(aviris_nc)
aoi <- project(vect(aoi_geojson), crs(aviris))
```

Crop first for speed, then mask to AOI boundary.

```{r}
aviris <- mask( # Mask raster by area of interest (AOI) boundary
  crop(aviris, aoi) # Crop to AOI before masking (more efficient apparently)
  [[grep("reflectance",names(aviris))]], # keep only reflectance bands for analysis (remove water vapour and aerosol bands); adjust if needed
  aoi, # AOI - read in as a separate vector file
  filename = "data/aviris_cropped.tif", # save masked raster to disk so R doesn't have to keep it in memory - more efficient
  overwrite = TRUE) # allow overwriting existing file if one already exists
```


# Vegetation indices 

Before we get into hyperspectral analyses, let's calculate some common vegetation indices that can be derived from the AVIRIS-NG data. These are essentially the way you would analyse traditional multispectral data that you get from existing satellite missions like Landsat, MODIS or Sentinal and can be a useful first step to explore the data and identify areas of interest.

First we select the bands we want to use for calculating vegetation indices. The specific band numbers will depend on the wavelength-to-band mapping for your AVIRIS-NG data, which you can find in the metadata/header of your NetCDF file. While each band covers a range of the electromagnetic spectrum, the example below assumes that the NIR band is around 860 nm, the RED band is around 660 nm, the GREEN band is around 550 nm, the SWIR band is around 1240 nm, and the BLUE band is around 450 nm. You would need to update these band numbers based on your specific dataset.

```{r}
band_nir <- 98L # see names(aviris)[98]
band_red <- 59L
band_green <- 36L
band_swir <- 174L
band_blue <- 15L 
```

## Extract RGB bands to plot a true colour image

By for the most common use of multispectral imagery is to create an RGB image, which is a true colour representation of the landscape. To do this, we need to extract the red, green and blue bands from our AVIRIS-NG data and assign them appropriate names for plotting.

```{r}
red <- aviris[[band_red]]
green <- aviris[[band_green]]
blue <- aviris[[band_blue]]
names(red) <- "red"
names(green) <- "green"
names(blue) <- "blue"

plotRGB(c(red, green, blue), stretch = "lin", main = "AVIRIS True Color Composite")
```

Or interactively...

```{r}
viewRGB(raster::stack(c(red, green, blue)), r = 1, g = 2, b = 3)
```


## Calculate vegetation indices

Here we calculate the Normalized Difference Vegetation Index (NDVI), the Normalized Difference Water Index (NDWI), and the Enhanced Vegetation Index (EVI) using the appropriate bands from our AVIRIS-NG data. These indices are commonly used in remote sensing to assess vegetation health, water content, and other ecological characteristics.

```{r}
# Extract NIR and SWIR bands for index calculations
nir <- aviris[[band_nir]]
swir <- aviris[[band_swir]]
names(nir) <- "nir"
names(swir) <- "swir"

# NDVI = (NIR - RED) / (NIR + RED)
ndvi <- (nir - red) / (nir + red)
names(ndvi) <- "NDVI"

# NDWI (Gao-style using NIR and SWIR) = (NIR - SWIR) / (NIR + SWIR)
ndwi <- (nir - swir) / (nir + swir)
names(ndwi) <- "NDWI"

# EVI requires a BLUE band, in addition to red and nir.
evi <- 2.5 * (nir - red) / (nir + 6 * red - 7.5 * blue + 1)
names(evi) <- "EVI"

veg_indices <- c(ndvi, evi, ndwi)
plot(veg_indices, nc = 3)
```

Again, these may be more interesting if viewed interactively... Here's EVI, but set to be slightly transparent (see `alpha.regions =`) so we can see the underlying landscape.

You'll see that areas with large shrubs tend to be highlighted with higher EVI.

```{r}
mapview(evi, alpha.regions = 0.4, map.types = "Esri.WorldImagery")
```

In fact, if you choose the right EVI threshold you can do a reasonable job of mapping the large shrubs and wetland areas.

```{r}
mapview(app(evi>0.3, as.integer), alpha.regions = 0.35, map.types = "Esri.WorldImagery")
```


## Point data

Now let's have a look at some point locations that represent where the two large `shrub` species (_Mimetes fimbriifolius_ and _Leucospermum conocarpodendron_) were observed, in addition to some `rock`, low vegetation (`lowveg`) and `path`s.

The points were collected from high resolution imagery in Google Earth and are stored in a KML file, which we will read into R and reproject to match the coordinate reference system of the AVIRIS data before plotting. Hover your curso over each point to see what cover class it represents.

```{r}
pts <- project(vect(points_file), crs(aviris))

mapview(pts, label = pts$Name, cex = 5, map.types = "Esri.WorldImagery")
```

Sticking with the "multispectral" and "vegetation index" approach, let's extract the EVI, NDVI and NDWI values at each of the point locations and plot them to see if there are any differences between the different cover classes.

```{r}
# Extract vegetation index values and selected AVIRIS bands to points.
terra::extract(veg_indices, pts) |>
  mutate(EndMember = pts$Name) |>
  na.omit() |>
  pivot_longer(cols = c("NDVI", "EVI", "NDWI"), names_to = "Index", values_to = "Value") |>
  ggplot(aes(x = EndMember, y = Value, fill = EndMember)) +
    geom_boxplot() +
    facet_wrap(~ Index, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Based on this, you would think you could just bin the NDVI raster into ranges of values to map the large shrubs, low vegetation and a combined class for rock and path.

```{r}
# Define a matrix of ranges and numericclass: [from, to, becomes_integer_ID]
# Values 1-33 -> 1, 33-66 -> 2, 66-100 -> 3
reclass_matrix <- matrix(c(0.7,  1,  1,
                           0.4, 0.7,  2,
                           0, 0.4, 3), 
                         ncol=3, byrow=TRUE)

# Apply the classification
# include.lowest=TRUE ensures the absolute minimum value is counted
rc <- classify(ndvi, reclass_matrix, include.lowest=TRUE)

plot(rc, main = "Classified by NDVI")
```

Or interactively...

```{r}
mapview(rc, col.regions = rainbow(3), alpha.regions = 0.35, map.types = "Esri.WorldImagery")
```

Not bad. The problem is that there is a lot of variation in NDVI values within the same cover class, and there is also overlap between the classes. This is where hyperspectral analyses can be really useful, as they can help to tease apart these differences by looking at the full spectrum of reflectance values rather than just a few indices.


# Imaging spectroscopy

The analysis presented so far is a bread and butter example of how one may process multispectral imagery (imagery that measures up to 20 or so broad bands of wavelengths across the electromagnetic spectrum) to extract information about vegetation. However, the real power of imaging spectroscopy (hyperspectral imagery) is that it provides a continuous spectrum of reflectance values across hundreds of narrow bands, which can be used to identify and differentiate between different materials and vegetation types based on their unique spectral signatures.

## Spectral signatures

First, let's inspect the spectral signatures of the different cover classes at our point locations. We can extract the reflectance values for all the bands at each point location and plot them to see if there are any differences in the spectral signatures between the different cover classes.

```{r, warnings = FALSE}
# Extract reflectance spectra at point locations to make a "spectral library" of the different cover classes
spectra_at_points <- terra::extract(aviris, pts) |>
  mutate(EndMember = pts$Name) |>
  na.omit()

# Plot spectral library
spectra_at_points |>
  group_by(EndMember, ID) |>
  pivot_longer(cols = starts_with("reflectance"), names_to = "Band", values_to = "Reflectance") |>
  mutate(Band = parse_number(str_split_i(Band, "=", 2))) |>
  ggplot(aes(x = Band, y = Reflectance, group = ID, color = EndMember)) +
  geom_line(alpha = 0.3) +
  stat_summary(aes(group = EndMember), fun = mean, geom = "line", linewidth = 1) +
  annotate(geom = "rect", xmin = 1350, xmax =  1420 , ymin = -Inf, ymax = Inf, 
            fill = "black", alpha = 0.7) + #1263-1562 nm 
  annotate(geom = "rect", xmin = 1785, xmax =  1975 , ymin = -Inf, ymax = Inf, 
            fill = "black", alpha = 0.7) + #1761-1958 nm
  theme_minimal() +
  ylim(0,0.5) +
  facet_wrap(~EndMember) + # facet_wrap(~Forest) +
  labs(title = "Spectral library of Silvermine Endmembers") +
  xlab("Wavelength (nm)")
```

Note the regions ~1350-1420 and ~1785-1975 nm are masked out in the plot above because they are heavily affected by absorption by water vapour in the atmosphere and tend to be very noisy, so they are often excluded from analyses of airborne or satellite imagery. You can see that there are differences in the spectral signatures between the different cover classes, which is promising for our ability to differentiate between them using hyperspectral data. For example, the large shrubs have a steep "red edge" due to higher absorption of red light (that plants want for photosynthesis) and higher reflectance in the NIR region (above 700 nm - that plants don't want, because it does not provide enough energy for photosynthesis, but does cause heating that can damage tissues) compared to the low vegetation and rock/path classes. 

## Multiple Endmember Spectral Mixture Analysis (MESMA)

One of the most common approaches to analyzing hyperspectral imagery is to use spectral mixture analysis (SMA - also called "spectral unmixing"), which assumes that the reflectance spectrum of each pixel in the image is a linear combination of the reflectance spectra of the different "endmembers" (i.e., the different cover classes) present in that pixel. The goal of SMA is to estimate the _fractional_ abundance of each endmember in each pixel, which can be used to create maps of the different cover classes. A key feature of the SMA approach is that it allows for the possibility that each pixel may contain a mixture of different endmembers, which is often the case in real-world landscapes. This is called the "mixed pixel problem", and becomes more relevant as ground sample distance (pixel size) increases or target endmembers (e.g. plant species) decrease in size.

One of the most common SMA approaches is Multiple Endmember Spectral Mixture Analysis (MESMA), which can be implemented in R using `library(RSToolbox)` or `library(luna)`. Here we'll use `luna`, which is being developed as a companion to the `terra` package, but with a focus on remote sensing. The basic idea of MESMA is to fit a linear model to each pixel in the image using the reflectance spectra of the different endmembers as predictors, and then use the coefficients of the model to estimate the fractional abundance of each endmember in that pixel. 

In `luna`, the analysis only takes a single representative of each endmember, so we will select a single spectrum for each endmember from our spectral library. A more robust approach would be to run multiple iterations of the analysis, randomly selecting from multiple spectra for each endmember, and summarising the outputs, but this is beyond the scope of this tutorial.

```{r}
# Select the first spectrum from each class for our spectral library
spectra_at_points[c(1,6,11,16),] |>
  remove_rownames() |>
  column_to_rownames(var = "EndMember") |>
  select(starts_with("reflectance")) -> spectral_library

# Run MESMA
mesma_result <- mesma(aviris, em = spectral_library[,], iterate = 400)

plot(mesma_result)
```

The output is a stack of rasters; one for each endmember (ranging from 0 to 1 to represent fractional cover) and another presenting the Root Mean Square Error (RMSE) of the model fit for each pixel. The RMSE raster can be used to assess the quality of the model fit, with lower values indicating a better fit. 

Again, we can visualize the results interactively using `mapview`. Here we will visualize the fractional cover of the `shrub` endmember, which is the one we are most interested in mapping.

```{r}
mapview::mapview(mesma_result$shrub, alpha.regions = 0.35, map.types = "Esri.WorldImagery")
```

Is this any better than EVI or NDVI? Let's compare the two approaches by plotting the fractional cover of the `shrub` endmember from MESMA against the EVI values for each pixel. 

```{r}
# Stack rasters and plot pairwise relationships
pairs(c(mesma_result$shrub, evi, ndvi))
```

While there's a fair correlation between them, there's also a fair difference between them and the correlation between the two vegetation indices is far stronger than between either and the MESMA estimate of shrub cover. The MESMA approach should able to better differentiate between the different cover classes, as it takes into account the full spectral information rather than just one or a few indices. This should be particularly useful in areas where there is a lot of variation in vegetation types and cover classes, as it allows for a more nuanced understanding of the landscape.

# Differentiating among our shrub species?

Our first goal is to determine if the two species do indeed have unique spectral signatures. One can do this at the leaf level (using leaf spectra collected with a field spectrometer with a leaf clip attachment), or by extracting points from the AVIRIS-NG data that represent pure pixels of each species. The latter is what we have done here, but it is important to note that the spectral signatures extracted from the AVIRIS-NG data may be influenced by factors such as canopy structure, background soil reflectance, and atmospheric effects, which can make it more difficult to differentiate between the two species compared to leaf-level spectra.

Let's get our points for each species, extract and compare.

```{r}
# Get points for each species
mpts <- vect("data/mimetes_canopy.geojson")
lpts <- vect("data/leucospermum_canopy.geojson")
cpts <- rbind(mpts, lpts) # combine into one vector object

# Extract spectra for each species
spectra_species <- terra::extract(aviris, cpts) |>
  mutate(Species = cpts$Name) |>
  na.omit()

# Plot spectral signatures for each species
spectra_species |>
  group_by(Species, ID) |>
  pivot_longer(cols = starts_with("reflectance"), names_to = "Band", values_to = "Reflectance") |>
  mutate(Band = parse_number(str_split_i(Band, "=", 2))) |>
  ggplot(aes(x = Band, y = Reflectance, group = ID, color = Species)) +
  geom_line(alpha = 0.3) +
  stat_summary(aes(group = Species), fun = mean, geom = "line", linewidth = 1) +
  annotate(geom = "rect", xmin = 1350, xmax =  1420 , ymin = -Inf, ymax = Inf, 
            fill = "black", alpha = 0.7) + #1263-1562 nm 
  annotate(geom = "rect", xmin = 1785, xmax =  1975 , ymin = -Inf, ymax = Inf, 
            fill = "black", alpha = 0.7) + #1761-1958 nm
  theme_minimal() +
  # ylim(0,0.5) +
  # facet_wrap(~Species) + # facet_wrap(~Forest) +
  labs(title = "Mimetes vs Leucospermum spectra") +
  xlab("Wavelength (nm)")
```

Pretty similar... Another approach is to do a Principal Component Analysis (PCA) on the spectra to see if the two species can be separated in the PCA space.

```{r}
# Do PCA
hyp.pca <- prcomp(spectra_species |> select(starts_with("reflectance")), center = TRUE, scale. = TRUE)
#screeplot(hyp.pca)

#autoplot(hyp.pca, data = mlib, colour = "Forest")

# Augment data with PCA results and original groupings
pca_data <- augment(hyp.pca, spectra_species$Species)

# Plot with ellipses
ggplot(pca_data, aes(x = .fittedPC1, y = .fittedPC2, color = Species)) +
  geom_point() +
  stat_ellipse() + # Adds 95% confidence ellipses
  theme_minimal() +
  labs(title = "PCA grouping by species",
       x = "PC1",
       y = "PC2")
```

